Notes on \\ Certified Projected Knowledge Compilation \\ \today

### 2023-03-27 doc/pmc-notes.tex
Initial thoughts on projected model counting

\newcommand{\varset}{X}
\newcommand{\exvarset}{Z}
\newcommand{\dependencyset}{{\cal D}}
\newcommand{\litset}{{\cal L}}
\newcommand{\ring}{{\cal R}}
\newcommand{\dset}{{\cal A}}
\newcommand{\rep}{\textbf{R}}
\newcommand{\radd}{+}
\newcommand{\rmul}{\times}
\newcommand{\addident}{\textbf{0}}
\newcommand{\mulident}{\textbf{1}}
\newcommand{\imply}{\Rightarrow}
\newcommand{\ifandonlyif}{\Leftrightarrow}
%\newcommand{\drational}{\mathbb{Q}_{2,5}}
\newcommand{\drational}{\textbf{Q}_{2,5}}

\newcommand{\assign}{\alpha}
\newcommand{\passign}{\rho}
\newcommand{\lassign}{\beta}
\newcommand{\uassign}{{\cal U}}
\newcommand{\modelset}{{\cal M}}

@@

\newcommand{\progname}[1]{\textsc{#1}}
\newcommand{\dfour}{\progname{D4}}
\newcommand{\Dfour}{\progname{D4}}
\newcommand{\cadical}{\progname{CaDiCal}}
\newcommand{\dtrim}{\progname{drat-trim}}

@@

\title{Certified Knowledge Compilation \\ with Application to Verified Model Counting}

\titlerunning{Certified Knowledge Compilation}

@@

\newcommand{\program}[1]{\textsc{#1}}
\newcommand{\lean}{Lean~4}

@@

\begin{abstract}

Computing many useful properties of Boolean formulas, such as their weighted or unweighted model count,
is intractable on general representations. It can become tractable when formulas are expressed in a
special form, such as the deterministic,
decomposable, negation normal form (d-DNNF)\@.
\emph{Knowledge compilation} is the process of converting a formula
into such a form. Unfortunately existing knowledge compilers provide no guarantee that their output correctly
represents the original formula, and therefore they cannot validate a model count, or any other computed value.

We present \emph{Partitioned-Operation Graphs} (POGs), a generalization
of the representations used by existing knowledge compilers.
We have designed  CPOG, a framework that can express proofs of equivalence between a
POG  and a Boolean formula in conjunctive normal form (CNF).

We have developed a program that generates POG representations from
the d-DNNF graphs produced by the state-of-the-art knowledge compiler
\dfour{}, as well as checkable CPOG proofs certifying that the output POGs
are equivalent to the input CNF formulas.  Our toolchain
for generating and verifying POGs scales to all but the largest d-DNNF
graphs produced by \dfour{} for formulas from a recent model counting
competition. Additionally, we have developed a formally verified CPOG
checker and model counter for POGs in the \lean{} proof assistant.
In doing so, we proved the soundness of our proof framework. These programs
comprise the first formally verified toolchain for weighted and unweighted
model counting.
\end{abstract}

\section{Introduction}

Given a Boolean formula $\phi$, modern Boolean satisfiability (SAT) solvers can
find an assignment satisfying $\phi$ or generate a proof that no
such assignment exists.  They have applications across a variety of
domains including computational mathematics, combinatorial
optimization, and the formal verification of hardware, software, and
security protocols.  Some applications, however, require going
beyond Boolean satisfiability.  For example, the {\em model
  counting problem} requires computing the number of satisfying
assignments of a formula, including in cases where there are far too many
to enumerate individually.  Model counting has
applications in artificial intelligence, computer security, and
statistical sampling.  There are also many useful extensions of standard model counting,
including {\em
  weighted model counting}, where a weight is defined for
each possible assignment, and the goal becomes to compute the sum of the weights
of the satisfying assignments.

Model counting is a challenging problem---more challenging than the
already NP-hard Boolean satisfiability.  Several
tractable variants of Boolean satisfiability, including 2-SAT, become
intractable when the goal is to count models and not just determine
satisfiability \cite{valiant:siam:1979}.  Nonetheless, a number of
model counters that scale to very large formulas have been developed, as
witnessed by the progress in recent model counting competitions.

One approach to model counting, known as {\em knowledge compilation},
transforms the formula into a structured form for which model counting
is straightforward.  For example, the {\em deterministic,
  decomposable, negation normal form} (d-DNNF) introduced by
Darwiche~\cite{darwiche:aaai:2002,darwiche:ecai:2004} represents a
Boolean formula as a directed acyclic graph, with terminal nodes
labeled by Boolean variables and their complements, and with each
nonterminal node labeled by a Boolean And or Or operation.  Restrictions
are placed on the structure of the graph such that a count of the
models can be computed by a single bottom-up traversal.
Kimmig et al.\ present a very general {\em
  algebraic model counting}~\cite{kimmig:jal:2017} framework describing
properties of Boolean functions that can be efficiently computed from
a d-DNNF representation.  These include standard and weighted model
counting, and much more.

One shortcoming of existing knowledge compilers is that they have no
way to validate that their generated results are correct, i.e., that
the compiled representation is logically equivalent to the original
formula.  By contrast, all modern SAT solvers can generate
checkable proofs when they encounter unsatisfiable formulas.  The
guarantee provided by a checkable certificate of correctness enables
users of SAT solvers to fully trust their results.  Experience has also
shown that proof production capabilites allow SAT solver developers to quickly
detect and diagnose bugs in their programs. This, in turn, has led
to more reliable SAT solvers.

This paper introduces \emph{Partitioned-Operation Graphs} (POGs),
a generalization of the representations produced by current knowledge
compilers. The CPOG (for ``certified'' POG) file format then
captures both the structure of a POG
and a checkable proof of its logical equivalence to a Boolean formula in
conjunctive normal form (CNF).  A CPOG
proof consists of a sequence of clause addition and deletion steps,
based on an extended resolution proof system~\cite{Tseitin:1983}.
We establish a set of conditions that, when satisified by a CPOG file, guarantee that it
encodes a well-formed POG and provides a valid equivalence proof.

\begin{figure}
\centering{
\begin{tikzpicture}[scale=0.17]
  \draw (8,10) rectangle (20,18);
  \node at (14,15.2) {Knowledge};
  \node at (14,12.8) {Compiler};

  \draw (28,10) rectangle (40,18);
  \node at (34,15.2) {Proof};
  \node at (34,12.8) {Generator};

  \draw (50,10) rectangle (62,18);
  \node at (56,15.2) {Proof};
  \node at (56,12.8) {Checker};

  \draw (50,0) rectangle (62,8);
  \node at (56,5.2) {Ring};
  \node at (56,2.8) {Evaluator};

  \draw[dashed] (48,-2) rectangle (64,22);
  \node at (56,20) {Trusted Code};

  %% CNF
  \draw[thick] (0,20) -- (44,20) -- (44,16) [-latex] -- (50,16);
  \draw[thick] (4,20) -- (4,16) [-latex] -- (8,16);
  \draw[thick] (24,20) -- (24,16) [-latex] -- (28,16);
  \node [left] at (-1,20) {\large {$\inputformula$}};
  \node [above] at (2,20) {\texttt{.cnf}};

  %% NNF
  \draw[thick] (20,12) [-latex] -- (28,12);
  \node [above] at (24,12) {\texttt{.ddnnf}};

  %% CPOG
  \draw[thick] (40,12) [-latex] -- (50,12);
  \draw[thick] (44,12) -- (44,4) [-latex] -- (50,4);
  \node [above] at (44,12) {\texttt{.cpog}};

  %% OK/Not
  \draw[thick] (62,14) [-latex] -- (66,14) ;
  \node [right] at (66,15.2) { OK /} ;
  \node [right] at (66,12.8) { Not OK} ;
  \draw[thick] (62,4) [-latex] -- (66,4) ;
  \node [right] at (66,4) {\large {$\textbf{R}(\inputformula)$}};

\end{tikzpicture}
}
\caption{Certifying toolchain.
  The output of a standard knowledge compiler is converted into a combined graph/proof (CPOG)
  which can be independently checked and evaluated}
\label{fig:chain}
\end{figure}

Figure~\ref{fig:chain} illustrates our certifying knowledge
compilation and model counting toolchain.  Starting with
input formula $\inputformula$, the \dfour{} knowledge compiler~\cite{lagniez:ijcai:2017}
generates a d-DNNF representation, and the {\em proof
  generator} uses this to generate a CPOG file.
The {\em proof checker} verifies the equivalence of the CNF and CPOG representations.
The {\em ring evaluator} computes
a standard or weighted model count from the POG\@ representation.
As the dashed box in Figure~\ref{fig:chain} indicates, this toolchain
moves the root of trust away from the complex and highly
optimized knowledge compiler to a relatively simple checker and
evaluator.  Importantly, the proof generator need not be
trusted---its errors will be caught by the proof checker.

To ensure soundness of the abstract CPOG proof system, as well as correctness of its
concrete implementation, we formally verified the proof system as well as versions
of the proof checker and ring evaluator in the \lean{} proof assistant~\cite{demoura:cade:2021}.
Running these two programs on a particular CPOG file gives assurance that the proof
and the model count are correct. Our experience with developing a formally verified
proof checker has shown that, even within the well-understood framework of extended
resolution, it can be challenging to formulate a full set of requirements that guarantee soundness.
In fact, subtle details in the partitioned sum rule were identified during this process.

We evaluate our toolchain using benchmark
formulas from the 2022 standard and weighted model competitions.  Our
tools handle all but the largest d-DNNF graphs generated
by \dfour{}.  We also measure the benefits of several optimizations as
well as the relative performance of the verified checker with one
designed for high performance and capacity.

\section{Related Work}

Generating proofs of unsatisfiability in SAT solvers has a long
tradition~\cite{ZhangMalik} and has become widely accepted due to the
formulation of clausal proof systems for which proofs can readily be
generated and efficiently checked
\cite{RAT,wetzler14_drattrim}.
A number of formally verified checkers have been developed within different verification frameworks~\cite{cruz-cade-2017,lrat,Lammich:20,Tan:2021}.
The associated proofs add clauses while preserving satisfiability until the empty clause is derived.
By contrast, our proofs construct a new propositional formula, and we must verify
%both
that it
%satisfies a set of rules and that it
is equivalent to the input formula.
This requires verifying additional proof steps, including clause deletion steps, and subtle invariants, as described in
Sections~\ref{sect:cpog} and \ref{section:formally-verified-toolchain}.

Fichte, Hecher, and Roland~\cite{fichte:sat:2022} devised the MICE
proof framework for model counting programs.  Their proof rules are
based on the algorithms commonly used by model counters.
They present a series of lemmas and a theorem stating
that their checking framework is sound.  They also present
experimental results for cases where they modify existing model
counters to generate proof traces, and where they generate proof
traces from d-DNNF graphs.
%% They have not formally verified their
%%proof framework or created a formally verified checker.
Our toolchain is not directly comparable to theirs---they directly
certify the numeric value produced by a model counter, while we
certify that the graph generated by a knowledge compiler is logically
equivalent to the input formula.  We can then validate a standard or
weighted count from that representation.
%% Their tool
%% chain can be applied to some programs that are pure model counters,
%%while ours can be applied to pure knowledge compilers.
Nonetheless,
we compare the performance of our toolchain to theirs as part of our
experimental evaluation in Appendix~\ref{app:experiments}.

We know of no other
work on certifying the output of a knowledge compiler.

\section{Logical Foundations}
\label{section:logical:foundations}

  Let $\varset$ denote a set of Boolean variables, and let $\assign$
  be an {\em assignment} of truth values to some subset of the
  variables, where $0$ denotes false and $1$ denotes true, i.e.,
  $\assign \colon \varset' \rightarrow \{0,1\}$ for some $\varset'
  \subseteq \varset$.  We say the assignment is {\em total} when it
  assigns a value to every variable ($\varset' = \varset$), and that
  it is {\em partial} otherwise.
  The set of all possible total assignments over
  $\varset$ is denoted $\uassign$.

For each variable $x \in \varset$,
  we define the {\em literals} $x$ and $\obar{x}$, where $\obar{x}$ is the
  negation of $x$. An
  assignment $\assign$ can be viewed as a set of literals, where
  we write $\lit \in \assign$ when $\lit = x$ and $\assign(x) = 1$ or when
  $\lit = \obar{x}$ and $\assign(x) = 0$.  We write the negation of literal $\lit$ as $\obar{\lit}$.  That is, $\obar{\lit} = \obar{x}$ when $\lit = x$ and
$\obar{\lit} = x$ when $\lit = \obar{x}$.

\begin{definition}[Boolean Formulas]
  The set of Boolean formulas is defined recursively.  Each
  formula $\phi$ has an associated {\em dependency set}
  $\dependencyset(\phi)  \subseteq \varset$, and a set of models $\modelset(\phi)$,
  consisting of total assignments that satisfy the formula:
  \begin{enumerate}
  \item Boolean constants $0$ and $1$ are Boolean formulas,
    with $\dependencyset(0) = \dependencyset(1) = \emptyset$, with $\modelset(0) = \emptyset$, and with $\modelset(1) = \uassign$.
  \item Variable $x$ is a Boolean formula, with $\dependencyset(x) = \{x\}$
    and $\modelset(x) = \{\assign \in \uassign | \assign(x)=1\}$.
  \item For formula $\phi$, its {\em negation}, written $\boolnot \phi$ is a Boolean formula,
    with $\dependencyset(\boolnot \phi) = \dependencyset(\phi)$ and $\modelset(\boolnot \phi) = \uassign - \modelset(\phi)$.
  \item For formulas $\phi_1, \phi_2, \ldots, \phi_k$, their {\em product} $\phi = \bigwedge_{1 \leq i \leq k} \phi_i$ is a Boolean formula, with
      $\dependencyset(\phi) = \bigcup_{1 \leq i \leq k} \dependencyset(\phi_i)$ and
      $\modelset(\phi) = \bigcap_{1 \leq i \leq k} \modelset(\phi_i)$.
  \item For formulas $\phi_1, \phi_2, \ldots, \phi_k$, their {\em sum} $\phi = \bigvee_{1 \leq i \leq k} \phi_i$ is a Boolean formula, with
      $\dependencyset(\phi) = \bigcup_{1 \leq i \leq k} \dependencyset(\phi_i)$ and
      $\modelset(\phi) = \bigcup_{1 \leq i \leq k} \modelset(\phi_i)$.
  \end{enumerate}
\label{def:boolean}
\end{definition}

  We highlight two special classes of Boolean formulas.  A formula is
  in {\em negation normal form} when negation is applied only to variables.  A
  formula is in {\em conjunctive normal form} (CNF) when i) it is in
  negation normal form, and ii) sum is applied only to literals.  A CNF
  formula can be represented as a set of {\em clauses}, each of which is a
  set of literals.  Each clause represents the sum of the
  literals, and the formula is the product of its clauses.  We use
  set notation to reference the clauses within a formula and the
  literals within a clause.  A clause consisting of a single literal is referred to as a {\em unit} clause and the literal as a {\em unit} literal.
This literal must be assigned value $1$ by any satisfying assignment of the formula.

\begin{definition}[Partitioned-Operation Formula]\label{def:partitioned-operation-formula}
  A {\em partitioned-operation formula}
 satisfies the following for all product and sum operations:
      \begin{enumerate}
      \item The arguments to each product must have disjoint dependency sets.  That is, operation
        $\bigwedge_{1 \leq i \leq k} \phi_i$ requires $\dependencyset(\phi_i) \cap \dependencyset(\phi_j) = \emptyset$ for $1 \leq i < j \leq k$.
      \item The arguments to each sum must have disjoint models.  That is, operation
        $\bigvee_{1 \leq i \leq k} \phi_i$ requires $\modelset(\phi_i) \cap \modelset(\phi_j) = \emptyset$ for $1 \leq i < j \leq k$.
      \end{enumerate}
\end{definition}
     We let $\pand$ and $\por$ denote the product and sum operations in a partitioned-operation formula.

  \section{Ring Evaluation of a Boolean Formula}

We propose a general framework for summarizing properties of Boolean
formulas along the lines of algebraic model counting~\cite{kimmig:jal:2017}.

\begin{definition}[Commutative Ring]
  A {\em commutative ring} $\ring$ is an algebraic structure
  $\langle \dset, \radd, \rmul, \addident, \mulident \rangle$,
  with elements in the set $\dset$ and with commutative and
  associative operations $\radd$ (addition) and $\rmul$ (multiplication),
  such that multiplication distributes
  over addition.  $\addident$ is the additive identity and $\mulident$ is
  the multiplicative identity.  Every element $a \in \dset$ has an
  {\em additive inverse} $-a$ such that $a + -a = \addident$.
\label{def:ring}
\end{definition}
We write $a - b$ as a shorthand for $a + -b$.

\begin{definition}[Ring Evaluation Problem]
\label{def:ring_evaluation}
  For commutative ring $\ring$, a {\em ring weight function} associates a value $w(x) \in \dset$ with
  every variable $x \in \varset$.  We then define $w(\obar{x}) \doteq \mulident-w(x)$.

  For Boolean formula $\phi$ and ring weight function $w$, the {\em ring evaluation problem} computes
  \begin{eqnarray}
    \rep(\phi, w) & = & \sum_{\alpha \in \modelset(\phi)} \;\; \prod_{\lit \in \alpha} w(\ell) \label{eqn:rep}
  \end{eqnarray}
  In this equation, sum \scalebox{0.8}{$\sum$} is computed using addition operation $\radd$, and product \scalebox{0.8}{$\prod$} is computed using multiplication operation $\rmul$.
\label{def:weight}
\end{definition}

Many important properties of Boolean formulas can be
expressed as ring evaluation problems.  The
(standard) {\em model counting} problem for formula $\phi$ requires determining $|\modelset(\phi)|$.
It can be cast as a ring evaluation problem by having $\radd$ and
$\rmul$ be addition and multiplication over rational numbers and using
weight function $w(x) = 1/2$ for every variable $x$.
Ring evaluation of formula $\phi$ gives the {\em density} of
the formula, i.e., the fraction of all possible total assignments that are
models.  For $n = |\varset|$, scaling the density by $2^n$
yields the number of models.  This formulation avoids the need for a ``smoothing'' operation,
in which redundant expressions are inserted into the formula~\cite{darwiche:jair:2002}.

The {\em weighted model counting}  problem is also defined over
rational numbers.  Some formulations  allow
independently assigning weights $W(x)$ and $W(\obar{x})$ for each variable $x$ and its complement, with the possibility that
$W(x) + W(\obar{x}) \not = 1$.
We can cast this as a
ring evaluation problem by letting $r(x) = W(x) + W(\obar{x})$,
performing ring evaluation with weight function $w(x) = W(x)/r(x)$ for each
variable $x$, and computing the weighted count
as $\rep(\phi, w)\; \rmul\; \prod_{x \in \varset} r(x)$.
Of course, this requires that $r(x) \not = 0$ for all $x \in \varset$.

The {\em function hashing problem} provides a test
of inequivalence for Boolean formulas.  That is, for $n = |\varset|$, let $\ring$ be a
finite  field with $|\dset| = m$ such that $m \geq 2 n$.  For each $x \in \varset$, chose a value from $\dset$ at random for $w(x)$.  Two formulas
$\phi_1$ and $\phi_2$ will clearly have $\rep(\phi_1, w) = \rep(\phi_2, w)$
if they are logically equivalent.
%, and if $\rep(\phi_1, w) \not = \rep(\phi_2, w)$, then they are clearly inequivalent.
If they are not equivalent, then
the probability that $\rep(\phi_1, w) \not = \rep(\phi_2, w)$ will be at
least $\left(1-\frac{1}{m}\right)^n \geq \left(1-\frac{1}{2n}\right)^n > 1/2$.
Function hashing can therefore be used as part of a
randomized algorithm for equivalence testing~\cite{blum:ipl:1980}.
For example, it can compare different runs on a single formula,
either from different compilers or from a single compiler with different configuration parameters.

%We compare ring evaluation to algebraic model counting in Section~\ref{sect:future}.

\section{Partitioned-Operation Graphs (POGs)}
\label{sect:pog}

Performing ring evaluation on an arbitrary Boolean formula could be intractable, but it is straightforward for a formula with partitioned operations:
\begin{proposition}
\label{prop:ring:eval}
Ring evaluation with operations $\boolnot$, $\pand$, and $\por$ satisfies the following for any weight function $w$:
\begin{eqnarray}
%% V1
\textstyle
\rep(\boolnot \phi,\; w) &=& \mulident - \rep(\phi, w) \label{eqn:ring:negation} \\
\textstyle
\scalebox{1.2}{\rep}\left(\Pand_{1 \leq i \leq k} \phi_i,\; w \right) &=& \prod_{1 \leq i \leq k} \rep(\phi_i, w) \label{eqn:ring:product} \\
\textstyle
\scalebox{1.2}{\rep}\left(\Por_{1 \leq i \leq k} \phi_i,\; w\right) &=& \sum_{1 \leq i \leq k} \rep(\phi_i, w) \label{eqn:ring:sum}
\end{eqnarray}
\end{proposition}
As is described in \cref{section:formally-verified-toolchain}, we have proved these three equations using \lean{}.

A {\em partitioned-operation graph} (POG) is a directed, acyclic graph
with nodes $N$ and edges $E \subseteq N \times N$.  We denote nodes with boldface symbols, such as $\nodeu$ and $\nodev$.
When $(\nodeu,\nodev) \in E$,
node $\nodev$ is said to be a {\em child} of node $\nodeu$.
The in- and out-degrees of node $\nodeu$ are defined as $\indegree(\nodeu) = | E \cap (N \times \{\nodeu\}) |$, and
$\outdegree(\nodeu) = | E \cap (\{\nodeu\} \times N) |$.
Node $\nodeu$ is said to be {\em terminal} if $\outdegree(\nodeu) = 0$.  A terminal node is labeled by a Boolean constant or variable.
Node $\nodeu$ is said to be {\em nonterminal} if $\outdegree(\nodeu) > 0$.
A nonterminal node is labeled by one of the three Boolean operations: $\boolnot$, $\pand$ or $\por$.
Node $\nodeu$ can be labeled with $\boolnot$ only if $\outdegree(\nodeu) = 1$.
It can be labeled with operation $\pand$ or $\por$ only if it satisfies the partitioning restriction for that operation.
Every POG has a designated {\em root node} $\noder$.

In essence, a POG represents a partitioned-operation
formula with a sharing of common subformulas.  Indeed, every node in the graph can be viewed as a partitioned-operation formula, and so we write
$\phi_{\nodeu}$ as the formula denoted by node $\nodeu$.
Each such formula has a set of models, and we write $\modelset(\nodeu)$ as a shorthand for $\modelset(\phi_{\nodeu})$.

We define the {\em size} of POG $P$, written $|P|$, to be the
the number of nodes labeled $\pand$ or $\por$ plus the number of edges from these nodes to their children.  Ring
evaluation of $P$ can be performed with at most $|P|$ ring
operations by traversing the graph from the terminal nodes up to
the root, computing a value $\rep(\nodeu, w)$ for each node $\nodeu$.
The final result is then $\rep(\noder, w)$.

POGs generalize the d-DNNF graphs in two ways:
\begin{itemize}
\item They allow negation of arbitrary nodes in the graph, not just
  variables.
  %, and so we allow it in the interest of generality.
%%  Negation could be useful in some contexts, such as when
%%  converting binary decision diagrams (BDDs) having complemented
%%  edges~\cite{brace-dac-1990,minato-dac-1990} into POGs.
\item They allow arbitrary arguments to a $\por$ operation, as long as
  these have disjoint models.  By contrast, each
  sum node $\nodeu$ in a d-DNNF must have two children $\nodeu_1$ and $\nodeu_0$, and for these there must be a {\em decision variable} $x$ such that
  any total assignment $\assign \in  \modelset(\nodeu_b)$ has $\assign(x)=b$, for $b \in \{0,1\}$.
%%Our generalization is provided to allow an
%%  encoding of Sentential Decision Diagrams (SDDs)~\cite{darwiche:ijcai:2011} as
%%  POGs.  This is discussed in Section~\ref{sect:future}, describing possible future research.
\end{itemize}
  These generalizations allow more flexibility in the POG
  representation while maintaining the ability to efficiently perform ring evaluation.

\section{Clausal Proof Framework}

We write (possibly subscripted) $\theta$ for
formulas encoded as clauses, possibly with extension variables.
We write (possibly subscripted)
$\phi$ for formulas
that use no extension variables.

A proof in our framework consists of a sequence of clause addition
and deletion steps, with each step preserving the set of solutions to
the original formula.
The status of the proof at any step is represented as
a set of {\em active} clauses $\theta$, i.e., those that
have been added but not yet deleted.
Our framework is based
on {\em extended} resolution~\cite{Tseitin:1983}, where proof
steps can introduce new {\em extension variables} encoding Boolean formulas over input and prior extension variables.
Let $Z$
denote the set of extension variables occuring in formula $\theta$.
Starting with  $\theta$ equal to input formula $\inputformula$,
the proof must maintain the invariant that
$\inputformula \ifandonlyif \exists Z\,\theta$.

Clauses can be added in two different ways.  One is when they serve as
the {\em defining clauses} for an extension variable.  This form
occurs only when defining $\pand$ and $\por$ operations, as is
described in Section~\ref{sect:cpog}.  Clauses can also be added or
deleted based on {\em implication redundancy}.  That is, when clause
$C$ satisfies $\theta \imply C$ for formula $\theta$, then it can either
be added to $\theta$ to create the formula $\theta \cup \{C\}$ or it can be deleted
from $\theta \cup \{C\}$ to create $\theta$.

We use {\em reverse unit propagation} (RUP) to certify
implication redundancy when adding or deleting
clauses~\cite{goldberg,vangelder08_verifying_rup_proofs}.
RUP
is the core rule supported by standard
proof checkers~\cite{RAT,wetzler14_drattrim} for propositional logic. It provides a simple and efficient
way to check a sequence of applications of the resolution proof rule~\cite{robinson-1965}.
Let $C = \{\lit_1, \lit_2, \ldots,\lit_p\}$ be a clause to be
proved redundant with respect to formula $\theta$.  Let $D_1, D_2, \ldots, D_k$ be a sequence of supporting
{\em antecedent} clauses, such that each $D_i$ is in $\theta$.
A RUP step
proves that $\bigwedge_{1\leq i \leq k} D_i \imply C$ by showing
that the combination of the antecedents plus the negation of $C$ leads
to a contradiction.  The negation of $C$ is the formula
$\overline{\lit}_1 \land \overline{\lit}_2 \land \cdots \land
\overline{\lit}_p$, having a CNF representation consisting of $p$ unit
clauses of the form $\obar{\lit}_i$ for $1 \leq i \leq p$.  A RUP
check processes the clauses of the antecedent in sequence, inferring
additional unit clauses.  In processing clause $D_i$, if all but one
of the literals in the clause is the negation of one of the
accumulated unit clauses, then we can add this literal to the
accumulated set.  That is, all but this literal have been falsified,
and so it must be set to true for the clause to be satisfied.  The
final step with clause $D_k$ must cause a contradiction, i.e., all of
its literals are falsified by the accumulated unit clauses.

Compared to the proofs of unsatisfiability generated by SAT solvers,
ours have important differences.  Most
significantly, each proof step must preserve the set of solutions with respect to the input variables;
our proofs must therefore justify both clause deletions and additions.
By contrast, an unsatisfiability proof need only guarantee that
no proof step causes a satisfiable set of clauses to become
unsatisfiable, and therefore it need only justify clause additions.

>>>>

\newcommand{\dvarset}{X}
\newcommand{\avarset}{Y}
\newcommand{\exvarset}{Z}
\newcommand{\dependencyset}{{\cal D}}
\newcommand{\litset}{{\cal L}}
\newcommand{\ring}{{\cal R}}
\newcommand{\dset}{{\cal A}}
\newcommand{\rep}{\textbf{R}}
\newcommand{\drep}{\textbf{R}_D}
\newcommand{\radd}{+}
\newcommand{\rmul}{\times}
\newcommand{\addident}{\textbf{0}}
\newcommand{\mulident}{\textbf{1}}
\newcommand{\imply}{\Rightarrow}
\newcommand{\ifandonlyif}{\Leftrightarrow}
%\newcommand{\drational}{\mathbb{Q}_{2,5}}
\newcommand{\drational}{\textbf{Q}_{2,5}}

\newcommand{\dassign}{\alpha}
\newcommand{\aassign}{\beta}
\newcommand{\acombine}{\!\cdot\!}
\newcommand{\passign}{\rho}
\newcommand{\lassign}{\beta}
\newcommand{\skolem}{\sigma}
\newcommand{\uassign}{{\cal U}}
\newcommand{\udassign}{\uassign_X}
\newcommand{\uaassign}{\uassign_Y}
\newcommand{\modelset}{{\cal M}}
\newcommand{\dmodelset}{{\cal M_{D}}}

@@

\newcommand{\progname}[1]{\textsc{#1}}
\newcommand{\dfour}{\progname{D4}}
\newcommand{\Dfour}{\progname{D4}}
\newcommand{\cadical}{\progname{CaDiCal}}
\newcommand{\dtrim}{\progname{drat-trim}}
\newcommand{\lean}{Lean~4}

@@

\title{Notes on \\ Certified Projected Knowledge Compilation \\ \today }

\titlerunning{Projected Knowledge Compilation}

@@

@@

\section{Notation}

\subsection{Data Model}

The set of Boolean variables is divided into two sets: {\em data} variables
$\dvarset$ and {\em auxiliary} variables $\avarset$.  We use separate
assignments for the two sets of variables:
$\dassign \colon \dvarset \rightarrow \{0,1\}$
 and
 $\aassign \colon \avarset \rightarrow \{0,1\}$, and the symbol $\acombine$ to combine the two assignments:
 $\dassign\acombine\aassign \colon \dvarset\cup\avarset \rightarrow \{0,1\}$.
We let $\udassign$, $\uaassign$, and $\uassign$ denote the set of all
possible assignments to the data variables, auxiliary variables, and
all variables, respectively.  From this, we can see that
$\uassign = \udassign \times \uaassign$.

For Boolean formula $\phi$ over the variables $\dvarset$ and
$\avarset$, the set of {\em models}, written $\modelset(\phi)$
consists of those assignments $\dassign \acombine \aassign$ that
satisfy the formula.  The set of {\em data models}, written $\dmodelset(\phi)$ is defined as:
\begin{eqnarray}
\dmodelset(\phi) & \doteq & \{ \dassign \in \udassign | \exists \aassign \in \uaassign \; \dassign\acombine\aassign \in \modelset(\phi) \} \label{eqn:dmodelset}
\end{eqnarray}

\subsection{Data Ring Evaluation}

  For commutative ring $\ring$, a {\em ring weight function} associates a value $w(x) \in \dset$ with
  every data variable $x \in \dvarset$.  We then define $w(\obar{x}) \doteq \mulident-w(x)$.

  For Boolean formula $\phi$ over data variables $\dvarset$ and auxiliary variables $\avarset$, and ring weight function $w$, the {\em data ring evaluation problem} computes
  \begin{eqnarray}
    \drep(\phi, w) & = & \sum_{\alpha \in \dmodelset(\phi)} \;\; \prod_{\lit \in \alpha} w(\ell) \label{eqn:rep}
  \end{eqnarray}

The goal of {\em projected knowledge compilation} is to generate a
representation of an input formula $\inputformula$ such that data ring
evaluation can be performed in polynomial time with respect to the
size of the representation.

\subsection{Skolem Formulas}

One method of instantiating the existential quantifier
(\ref{eqn:dmodelset}) is to provide formulas that compute an
assignment $\aassign$ for each possible $\dassign$.  For $y \in
\avarset$, define a {\em Skolem formula} $\skolem_y$ as a Boolean
formula over $\dvarset$.  It implicitly defines a function
$\skolem_y \colon \udassign \rightarrow \{0,1\}$.  Combining these functions for all $y \in
\avarset$ then defines a function $\skolem \colon \udassign
\rightarrow \uaassign$, where $\skolem(\dassign)$ gives the assignment
$\aassign$ such that $\aassign(y) = \skolem_y(\dassign)$.

A {\em valid Skolemization} of formula $\phi$ is then a set of Skolem formulas $\skolem_y$
such that every data assignment $\dassign\in \dmodelset(\phi)$ is extended by $\skolem$ to become a model of $\phi$, i.e.,
$\dassign \acombine \skolem(\dassign) \in \modelset(\phi)$.

\section{Example Formula}

Consider an encoding of the formula $(x_1 \land x_2) \lor x_3$ using
encoding variable $y$.  A full {\em Tseitin} encoding of $y
\leftrightarrow (x_1 \land x_2)$ requires three clauses, but we can
use a Plaisted-Greenbaum encoding in this context, encoding just $y
\rightarrow (x_1 \land x_2)$, represented by the clauses
$x_1 \lor \obar{y}$ and $x_2 \lor \obar{y}$.  We then encode the disujunction
with the clause $x_3 \lor y$.  The DIMACS representation is then as follows (using variable 4 to represent $y$):

\begin{center}
\begin{tabular}{lll}
\toprule
\makebox[5mm]{ID} & \makebox[15mm]{Clauses} & \\
\midrule
1 & \texttt{1 -4} & \texttt{0} \\
2 & \texttt{2 -4} & \texttt{0} \\
3 & \texttt{3  4} & \texttt{0}\\
\bottomrule
\end{tabular}
\end{center}

\begin{figure}
  \begin{minipage}{0.48\textwidth}
    (A) POG representation of $\inputformula$ \\[1.2ex]
    \input{dd/eg-preproj}
  \end{minipage}
  \begin{minipage}{0.48\textwidth}
  \end{minipage}

  \begin{minipage}{0.48\textwidth}
    (B) After projecting variable $y$ \\[1.2ex]
    \input{dd/eg-proj}
  \end{minipage}
  \begin{minipage}{0.48\textwidth}
    (C) Replace $y$ by Skolem formula $x_1 \land x_2$ \\[1.2ex]
    \input{dd/eg-proj-skolemize}
  \end{minipage}
  \caption{POG Representation of formula (A), and two ways of representing projections (B) and (C).
  A black dot on an edge denotes Boolean negation.}
  \label{fig:eg-proj}
\end{figure}

Figure \ref{fig:eg-proj}(A) shows the POG representation of the
example formula, derived from the d-DNNF representation produced by
\dfour{}.  We can see that \dfour{} split on variable $y$ at the top level, with the representation of
$\obar{y} \land \inputformula$ on the left and 
$y \land \inputformula$ on the right.

Figure \ref{fig:eg-proj}(B) shows the result of a method for projecting out variable
$y$ by computing the intersection of the sets of data models for
$\aassign(y) = 0$ and $\aassign(y) = 1$.  For this example, these are
given by the formula $x_1 \land x_2 \land x_3$, represented by POG
node $\nodep_5$.  Using DeMorgan's Laws, we can use a sum node to
``subtract'' these models from those given by
$\simplify{\inputformula}{\obar{y}}$, so that the top-level sum
$\nodes_8$ satisfies the disjoint-model requirement.

Figure \ref{fig:eg-proj}(C) shows the result of replacing $y$ by the
Skolem formula
$\skolem_y = x_1 \land x_2$.  In that case node
$\nodep_6$ of (A) becomes simply $x_1 \land x_2$, and this can be used
to replace $\obar{y}$ as a child of node $\nodep_5$.

Observe that both POGs (B) and (C) use an exclusion term so that the formula
on the left encodes those assignments satisfying $x_3 \land \neg (x_1 \land x_2)$.

\section{Research Questions}

\begin{itemize}
\item Algorithm to generate projection of POG over set of auxiliary variables $\avarset$
  \begin{itemize}
  \item Assume have clausal representation $\gamma_{\nodeu}$ for each node $\nodeu$.
  \item Assume can call \dfour{} as subroutine to derive POG representing shared models (could blow up)
  \item Assume POG derived from \dfour{} in NNF
  \end{itemize}
\item Algorithm to find Skolem formulas
  \begin{itemize}
  \item Easy if $y$ encoded as Tseitin variable
  \item Probably OK if $y$ encoded using some variant of Greenbaum-Plaisted
  \item Unclear how to solve in general
  \end{itemize}
\item Algorithm for forward implication proof
  \begin{itemize}
  \item Can use existing proof infrastructure
  \item Use SAT solver to generate proofs of mutual exclusion and implication
  \end{itemize}
\item Reverse implication proof
  \begin{itemize}
  \item Need new proof rule(s).  Projected formula allows arbitrary assignment $\aassign$.
  \item Skolem formulas might be helpful
  \item Need to show that these form valid Skolemization
  \end{itemize}
\end{itemize}
